Automatically Splitting a Two-Stage Lambda Calculus

### 2015-07-11 paper/intro.tex
fiddling with figures

While simple forms of frequency reduction include standard compiler
optimizations such as loop hoisting and common subexpression
elimination, J{\o}rring and Scherlis proposed the more general
transformation of {\em splitting} a program into multiple passes
(called pass separation in \,\cite{JS86-staging}). In our example,
splitting transforms the function $f$ into two passes $f_1$ and $f_2$
such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
evaluating the first pass on $x$, and using the result $z$ to evaluate
the second pass on each $y_i$.
%
\begin{lstlisting}
fun @$F_\textrm{multipass}(x, y_1, y_2, \ldots, y_m)$@ = 
  let @$z = f_1(x)$@
  in @$f_2(z, y_1);  f_2(z, y_2); \ldots; f_2(z, y_m)$@
\end{lstlisting}
%
The key difference between splitting and partial evaluation (or
metaprogramming) is that the former can be performed without access to
the first argument $x$; $F_\textrm{multipass}$ works for any $x$,
while $F_v$ is defined only for $x=v$.  Therefore, unlike partial
evaluation, splitting is a static program transformation
(``metastatic'' in parial evaluation terms) and does not require
dynamic code generation.

@@

Prior work on partial evaluation and metaprogramming has demonstrated
automatic application of these techniques on higher order functional
languages. In contrast, automatic splitting transforms have been
limited to simple languages that do not support recursion or
first-class
functions\,\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}. In
this paper, we extend this work by presenting a splitting algorithm
for a typed lambda calculus.

As our starting point, we consider an explicitly-two-staged, typed
lambda calculus \lang\ in the style of Davies\,\cite{davies96}, with a
$\fut$ modality denoting computation in a later stage
(\ref{sec:semantics}). We then present a splitting algorithm that
transforms \lang\ program into two passes corresponding to the stages
of computation in that program (\ref{sec:splitting}). These passes are
monostaged programs, expressed in a conventional functional language.

The most interesting case involves splitting a recursive mixed-stage
function (like $f$ in the example). This yields a first pass function
$f_1$ which computes from its first-stage argument $x$ a
\emph{boundary} data structure encoding the interstage communication
in $f(x,-)$. When $f$ recurs on $x$, this boundary structure itself
has a recursive type.  The second pass is a function $f_2$ which takes
the boundary data structure $f_1(x)$ and a second-stage argument
$y_i$, and completes the computation $f(x,y_i)$, traversing the
boundary structure in light of the particular argument $y_i$.

We have implemented this splitting algorithm
(\ref{sec:implementation}) and applied it to a number of staged
programs (\ref{sec:examples}) ranging from straight-line arithmetic
operations to recursive and higher-order functions. In the case of the
quickselect algorithm, which we discuss next, the split code is in
fact asymptotically faster over many $y_i$.

>>>>

While simple forms of frequency reduction include standard compiler
optimizations such as loop hoisting and common subexpression
elimination, J{\o}rring and Scherlis proposed the more general
transformation of {\em splitting} a program into multiple subfunctions
(called pass separation in \,\cite{JS86-staging}). In our example,
splitting transforms the function $f$ into two others $f_1$ and $f_2$
such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
evaluating the first function on $x$, and using the result $z$ to evaluate
the second function on each $y_i$.
%
\begin{lstlisting}
fun @$F_\textrm{multipass}(x, y_1, y_2, \ldots, y_m)$@ = 
  let @$z = f_1(x)$@
  in @$f_2(z, y_1);  f_2(z, y_2); \ldots; f_2(z, y_m)$@
\end{lstlisting}
%
The key difference between splitting and partial evaluation (or
metaprogramming) is that the former can be performed without access to
the first argument $x$; $F_\textrm{multipass}$ works for any $x$,
while $F_v$ is defined only for $x=v$.  Therefore, unlike partial
evaluation, splitting is a static program transformation
(``metastatic'' in partial evaluation terms) and does not require
dynamic code generation.

@@

Prior work on partial evaluation and metaprogramming has demonstrated
automatic application of these techniques on higher order functional
languages. In contrast, automatic splitting transforms have been
limited to simple languages\,\cite{knoblock96,Proudfoot:2001,Foley:2011,He:2014}.

In this paper, we present a splitting algorithm for \lang, 
a two-staged typed lambda calculus in the style of Davies\,\cite{davies96},
with support for recursion and first-class functions.
Like Davies, \lang\ uses a $\fut$ modality to denote computation in the second stage,
but to aid splitting we also add a $\curr$ modality to denote purely first-stage computations. 
The dynamic semantics (\ref{sec:semantics}) of \lang\ are that of Davies, 
modified to provide an eager behavior between stages, which is more intuitive in the context of splitting.
We then prove the correctness of our splitting algorithm (\ref{sec:splitting}) for \lang\ with respect to the semantics.
Finally, we discuss our implementation of this splitting algorithm
(\ref{sec:implementation}) and demonstrate its power and behavior
for a number of staged programs ranging from straight-line arithmetic
operations to recursive and higher-order functions (\ref{sec:examples}).

We demonstrate that splitting a recursive mixed-stage $f$
yields an $f_1$ which computes a recursive data structure
and an $f_2$ which traverses that structure
in light of information available at the second stage.
In the case of the quickselect algorithm, which we discuss next, 
the split code executes asymptotically faster than a unstaged evaluation of $f$.

### 2015-07-11 paper/paper.tex
various edits
attempting to lift some more discussion into 4.0
related work
fiddling with figures

\usepackage{cleveref}
\AtBeginDocument{\renewcommand\ref\cref}
\crefname{section}{Section}{Sections}
\crefname{figure}{Figure}{Figures}
\crefname{theorem}{Theorem}{Theorems}

@@

\title{Automatically Splitting a Staged Lambda Calculus}
\subtitle{(Supplementary Material in Appendix)}

\authorinfo{Anon A. Anon}%Nicolas Feltman \and Carlo Angiuli \and Umut A. Acar \and Kayvon Fatahalian}
           {Anonymous Institute}%Carnegie Mellon University}
           {anon@anon.com}%\{nfeltman,cangiuli,umut,kayvonf\}@cs.cmu.edu}

>>>>

\usepackage{cleveref}
\AtBeginDocument{\renewcommand\ref\cref}
\crefname{section}{Section}{Sections}
\crefname{figure}{Figure}{Figures}
\crefname{theorem}{Theorem}{Theorems}
\crefname{lemma}{Lemma}{Lemmas}

@@

\title{Automatically Splitting a Two-Stage Lambda Calculus}
\subtitle{(Supplementary Material in Appendix)}

\authorinfo{~}%Nicolas Feltman \and Carlo Angiuli \and Umut A. Acar \and Kayvon Fatahalian}
           {~}%Carnegie Mellon University}
           {~}%\{nfeltman,cangiuli,umut,kayvonf\}@cs.cmu.edu}

### 2015-07-11 paper/semantics.tex
fiddling with figures

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
whereas our applications feel more natural with a eager interpretation of {\tt next}.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

@@

The statics and dynamics of \lang\ are related by a type safety theorems
in figure \ref{fig:typeSafety}, again annotated by world.
In all cases, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second stage binders under which we are evaluating.

>>>>

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated 
when partial values are substituted for a variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
whereas our applications feel more natural with a eager interpretation of {\tt next}.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a \emph{hoisting step}.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a substitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

@@

The statics and dynamics of \lang\ are related by the type safety theorems
in \ref{fig:typeSafety}, again annotated by world.
In all cases, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second stage binders under which we are evaluating.

### 2015-07-11 paper/splitting.tex
fiddling with figures

The goal of a stage splitting translation 
is to a send program $\mathcal P$ in a multistage language
to an equivalent form $\mathcal P'$ where the stages are seperated at the top level.
By equivalent, we mean that $\mathcal P$ and $\mathcal P'$ should produce 
the same answer under their respective semantics.

@@

In order to prove that \ref{lem:twoCorrect},
we will need to state analogous version for steps at \bbonem,
since the various kinds multistage term in \lang\ are mutually dependent.
Thus, this section proceeds by covering the definition of splitting at \bbonem, 
starting with the value form at that world.

@@

To provide intuition about the behavior of partial value splitting, consider the
following partial value:
\begin{lstlisting}
1` (next{`2`y`1`}, (`3`gr{`1`injL 7`3`}`1`, next{`2`y`1`}))`
\end{lstlisting}
To construct the value $i$ representing its first-stage components,
splitting first redacts all second-stage (blue) parts, along with the
surrounding \texttt{next} annotations. 
The resulting ``holes'' in the term are replaced with
unit values.
%\begin{lstlisting}
%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
%\end{lstlisting}
%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
\begin{lstlisting}
1`((), (3`gr{`1`injL 7`3`}`,()))`
\end{lstlisting}
Finally, partial value splitting drops \texttt{gr} annotations yielding:
\begin{lstlisting}
1`((), (injL 7,()))`
\end{lstlisting}
To construct the residual $q$ (corresponding to second-stage computations) partial value splitting redacts all \texttt{gr} blocks
(replacing them with unit) and \texttt{next} annotations:
\begin{lstlisting}
2`(y, ((),y))`
\end{lstlisting}
A precise definition of the partial value splitting relation is given in
\ref{fig:termSplit}.

>>>>

The goal of a stage splitting translation 
is to send a program $\mathcal P$ in a multistage language
to an equivalent form $\mathcal P'$ where the stages are seperated at the top level.
By equivalent, we mean that $\mathcal P$ and $\mathcal P'$ should produce 
the same answer under their respective semantics.

@@

In order to prove \ref{lem:twoCorrect},
we will need to state analogous version for steps at \bbonem,
since the various kinds of multistage term in \lang\ are mutually dependent.
Thus, this section proceeds by covering the definition of splitting at \bbonem, 
starting with the value form at that world.

@@

To provide intuition about the behavior of partial value splitting, consider the
following partial value:
\begin{lstlisting}
1` (next{`2`y`1`}, (`3`gr{`1`injL 7`3`}`1`, next{`2`y`1`}))`
\end{lstlisting}
To construct the value $i$ representing its first-stage components,
splitting first redacts all second-stage (blue) parts, along with the
surrounding \texttt{next} annotations. 
The resulting ``holes'' in the term are replaced with
unit values.
%\begin{lstlisting}
%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
%\end{lstlisting}
%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
\begin{lstlisting}
1`((), (3`gr{`1`injL 7`3`}`,()))`
\end{lstlisting}
Finally, partial value splitting drops \texttt{gr} annotations, yielding:
\begin{lstlisting}
1`((), (injL 7,()))`
\end{lstlisting}
To construct the residual $q$ (corresponding to second-stage computations) partial value splitting redacts all \texttt{gr} blocks
(replacing them with unit) and \texttt{next} annotations:
\begin{lstlisting}
2`(y, ((),y))`
\end{lstlisting}
A precise definition of the partial value splitting relation is given in
\ref{fig:termSplit}.
